The Impact of Model Assumptions on RES Drought Analysis

### 2024-11-13 Article/main.tex
Update Article

\section{Introduction}

@@

\begin{comment}
The C3S-E national dataset assumes RES farms exist at every grid point, whereas for the C3S-E Gridded and Atlite datasets, the location of each RES farm is applied. \textcolor{magenta}{Do you think you could add somehow that the C3S-E datasets assume specific wind and PV conversion models? Perhaps this should not be in the introduction, but since farms are mentioned I would include the model as well.}
\end{comment}

This study aims to quantify the occurrence of RES drought events in Ireland for two energy scenarios. The first represents the current energy mix, predominantly driven by wind, while the second is based on projected energy scenarios, featuring a more balanced share between wind and PV. A consistent methodology is used to identify RES droughts for each of the three datasets, and differences in the frequency, the duration and the seasonality are presented.

\begin{comment}
	 \textcolor{magenta}{Did we say we are avoiding temporal references in scenarios? I get that ignoring what they represent is not the way to go but the line is very fine I would say}
\end{comment}

The datasets used in this study are detailed in section~\ref{sec:data}, which describes their sources, characteristics, and relevance for evaluating RES droughts. Section~\ref{sec:methods} outlines the energy models used to simulate wind and PV generation and provides the methodology for defining and identifying RES drought events, including the thresholds and metrics applied. In section~\ref{sec:analysis}, an in-depth analysis of RES drought occurrences is presented for the two energy scenarios, highlighting key differences across the datasets and the effects of varying energy capacities. Finally, section~\ref{sec:DiscConc} offers a discussion of the results in the context of energy reliability and future planning, followed by the main conclusions and recommendations for further research.

\section{Data}
\label{sec:data}

To construct and validate the models, this study uses publicly available datasets from EirGrid, the Republic of Ireland's transmission system operator, and the ERA5 reanalysis dataset. These datasets provide necessary inputs for estimating the CF for wind and PV energy generation.

\subsection{EirGrid}
\label{sec:eirgrid}

EirGrid provides detailed datasets for all wind and PV farms in the Republic of Ireland and Northern Ireland (collectively referred to as Ireland) from 1990 to the present~\cite{eirgrid}. These datasets include information on each farm’s installed capacity, name, and connection date. To improve these datasets, the GPS coordinates for each farm were manually determined using online search.

EirGrid also provides a wide range of energy system variables for wind (2014-present) and PV (2018-present) energy generation. For our study, we use generation which refers to the actual amount of energy produced by the wind or PV farms. This is then combined with the installed capacity to compute the CF. \textcolor{cyan}{Change}

\begin{comment}
boris: we use availability which indicates the potential energy output if the farms were operating at full capacity 
\end{comment}

@@

The Copernicus Climate Change Service (\textcolor{cyan}{I have defined C3S-E before, should I use it here ?} \textcolor{magenta}{I am not against introducing some acronyms twice, same as for some references being in the introduction and data, but I would say there is no right or wrong way necessarily} )has developed a renewable energy dataset for Europe~\cite{dubus2023energy}, using ERA5 atmospheric variables and weather-to-energy models. This dataset provides hourly CF for wind and PV energy from 1979 to the present. The data are available on the same grid as the ERA5 data, which has a horizontal resolution of 0.25°. Data are also available for download at two aggregated scales: regional (NUTS 2) and national.

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and actual trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: global horizontal irradiance (referred to here as $ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles~\cite{saintdrenan2018solar}. \textcolor{cyan}{I added a sentence to explain the reason of the Vestas choice} \textcolor{magenta}{So, basically what they are saying is that they are looking at a realistic turbine for the future, makes sense. Not necessarily linked to the article, but I wonder how realistic this turbine choice is for other countries. We know it is not for Ireland, but I am curious about the rest of Europe}

\section{Methods}
\label{sec:methods}

@@

\subsection{C3S-E N}
\label{sec:C3SEN}

@@

\subsection{C3S-E G}
\label{sec:C3SEG}

For the second dataset, we use the gridded dataset from C3S-E to create a CF time series which is representative of the actual RES farm distribution. Using the farms' coordinates, we identify the nearest grid point on the C3S-E grid for each farm. We retrieve the CF values from the C3S-E dataset for its corresponding grid point. We then average the CF for each farm using the installed capacity as weights. This process results in a CF time series for each renewable energy source (wind and PV), which is representative of the location of the RES farms in Ireland.

@@

For wind power estimation, Atlite uses the wind speed at 100 metres ($u_{100}$, $v_{100}$). We provide Atlite with a wind turbine power curve generated by Renewables.ninja: Enercon E112.4500 with 0.30w smoothing~\cite{staffell2016wake} at a hub height of 100 metres, which it uses when converting wind speed to CF.

For PV power estimation, Atlite uses four irradiance variables ($ssr$, $ssrd$, $tisr$, and $fdir$) as well as the air temperature ($t2m$). We use a single PV model~\cite{beyer2004pv}, with the Kaneka Hybrid panel option. The azimuth angle is fixed at 180\textdegree (due south) and we select the optimal tilt angle option.

@@

This study investigates wind and PV CF using three models, resulting in six CF time series with the installed capacity at the end of 2023. Additionally, we analyse the combined CF of wind and PV for each model by averaging the two sources, weighted by their capacities (6.7 GW for wind, 0.6 GW for PV). This approach gives us a total of nine CF time series for the 2023 scenario.

Given the relatively low PV capacity in Ireland in 2023 and the projected increase by 2030 (expected to increase from 0.6 GW to 8 GW), we also explore a future scenario with higher PV capacity. We generate additional CF time series reflecting the projected PV capacity in 2030 for both the Atlite and C3S-E G models, based on a roadmap published by EirGrid~\cite{eirgrid2023future}.

The wind capacity is projected to increase to 9 GW in 2030, however as the spatial distribution is not expected to significantly change we use the same CF time series. The combination of wind and PV CF is calculated once again using the new PV time series, and by averaging with the future projected capacities of RES farms. In total, we evaluate fourteen CF time series across different scenarios and models, as summarised in Table~\ref{tab:datasets}.

\begin{table}[h!]
	\centering
	\begin{tabular}{l|ccc|ccc|}
		\cline{2-7}
		& \multicolumn{3}{c|}{\textbf{Current}}                                   & \multicolumn{3}{c|}{\textbf{Projected}}                                   \\ \cline{2-7} 
	    & \multicolumn{1}{c|}{Atlite} & \multicolumn{1}{c|}{C3S-E G} & C3S-E N & \multicolumn{1}{c|}{Atlite} & \multicolumn{1}{c|}{C3S-E G} & C3S-E N \\ \hline
		\multicolumn{1}{|l|}{\textbf{PV}}      & \multicolumn{1}{c|}{x}      & \multicolumn{1}{c|}{x}       & x       & \multicolumn{1}{c|}{x}      & \multicolumn{1}{c|}{x}       &         \\
		\multicolumn{1}{|l|}{\textbf{Wind}}    & \multicolumn{1}{c|}{x}      & \multicolumn{1}{c|}{x}       & x       & \multicolumn{1}{c|}{}       & \multicolumn{1}{c|}{}        &         \\
		\multicolumn{1}{|l|}{\textbf{Combine}} & \multicolumn{1}{c|}{x}      & \multicolumn{1}{c|}{x}       & x       & \multicolumn{1}{c|}{x}      & \multicolumn{1}{c|}{x}       & x       \\ \hline
	\end{tabular}
	\caption{Summary of the time series compared in this study}
	\label{tab:datasets}
\end{table}

\subsection{RES drought definition}
\label{sec:res_drought}

In this study, we define a RES drought event as occurring when the 24-hour moving average of CF remains below a fixed threshold of 0.1 for a period of longer than 24 hours. Using the moving average captures fewer, but longer-lasting events than using the regular time series.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{data_methods/find_res_droughts}
	\caption{Time series of CF (red) and its 24-hour moving average (green) from the 7th to the 15th of July 2021. CF threshold (dashed black). The grey bar shows the period identified as a wind drought.}
	\label{fig:find_res_droughts}
\end{figure}

Fig.~\ref{fig:find_res_droughts} illustrates the moving average  identification method. When using the moving average method, this CF time series results in one event being identified, which lasts 161 hours. In contrast, using the regular CF series would have resulted in three separate events being identified, with durations of 36, 35, and 40 hours.

\section{Results}
\label{sec:results}

@@

\subsubsection{Wind}
\label{sec:wind_verification}

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve} a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provide four associated smoothed power curves. The smoothing is done by using a Gaussian filter with different standard deviations that depend on the wind speed. We generated a wind CF time series for Ireland for each of these wind turbine power curve options. The performance of each wind turbine power curve was calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$ filter.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification/power_curve}
	\caption{a/ Power curves of the Enercon E112.4500 with a 0.3 filter used by Atlite (blue) and the Vestas V136/3450 used by C3S-E (orange) b/ Histograms of wind CF for Ireland from Atlite (blue), C3S-E (orange) and EirGrid (shaded)}
	\label{fig:power_curve}
\end{figure}

The smoothing of the wind turbine power curve represents the wake effect, which is important when modelling wind energy on larger scales. The histogram in Fig.~\ref{fig:power_curve}b shows that the C3S-E power curve tends to underestimate low CF values and overestimate higher ones, whereas the smoothed Atlite power curve more closely follows the recorded wind data from EirGrid.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification/wind_verification_contour}
	\caption{Wind CF density plot of the observed EirGrid CF (vertical axes) and modelled (horizontal axes) CF data for the Atlite  (left panel), C3S-E G (central panel) and C3S-E N (right panel) models}
	\label{fig:wind_verification_contour}
\end{figure}

Fig.~\ref{fig:bar_number_events_verification_wind} presents results for the average annual number of wind drought events during the 2014 to 2023 validation period. The figure shows that Atlite shows the best agreement overall with the observed frequency and duration of wind drought events.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification/bar_number_events_verification_wind}
	\caption{Average annual number of wind drought events for EirGrid (black), Atlite (red), C3S-E G (blue), and C3S-E N (purple) / Change plot}
	\label{fig:bar_number_events_verification_wind}
\end{figure}

\subsubsection{PV}
\label{sec:pv_verification}

The C3S-E datasets calculate PV generation multiple times using the same model with varying azimuth and tilt angles. The results are then aggregated based on a statistical distribution of module angles. Similarly to the wind CF verification, the Atlite model allows the user to select from one of three different PV panel characteristics. Following the same methodology as in the previous section, we compare the three available models using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, we selected the Breyer PV panel model \cite{beyer2004pv}, specifically using the Kaneka Hybrid panel option.

Figure \ref{fig:solar_verification_contour} shows that the three datasets have similar correlation when compared to the EirGrid data, which is further supported by the results shown in Table~\ref{tab:pv_skill_scores}.

\begin{table}[!ht]
	\centering
	\begin{tabular}{l|lll|}
	\cline{2-4}
	& \textbf{Atlite} & \textbf{C3S-E G} & \textbf{C3S-E N} \\ \hline
	\multicolumn{1}{|l|}{\textbf{CC}}   & 0.921           & 0.931            & 0.913            \\ \hline
	\multicolumn{1}{|l|}{\textbf{RMSE}} & 0.070           & 0.065            & 0.100            \\ \hline
	\multicolumn{1}{|l|}{\textbf{MBE}}   & 0.002           & -0.012           & -0.013           \\ \hline
	\end{tabular}
	\caption{Skill scores for wind CF for the three datasets compared to EirGrid data}
	\label{tab:pv_skill_scores}
\end{table}

\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{verification/solar_verification_contour}
	\caption{PV 2D density plot of the observed EirGrid (vertical axes) and modelled (horizontal axes) CF series for the Atlite  (left panel), C3S-E G (central panel) and C3S-E N (right panel) models}	
	\label{fig:solar_verification_contour}
\end{figure}

Fig.~\ref{fig:bar_number_events_verification_pv} shows results for PV drought events. The figure reveals a lack of significant agreement between the three datasets and the EirGrid data. The main challenge in validating PV stems from the recent installation of a large proportion of the PV installed capacity in Ireland. PV Generation data suffers from large uncertainties for the first few months after each PV farm is connected, making it difficult to know what the actual CF for PV was. Given that PV farms installed in 2023 represent more than 60\% of the total capacity, this significantly impacts our ability to perform vigorous verification for PV.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification/bar_number_events_verification_pv}
	\caption{Average number of PV drought events for EirGrid (black), Atlite (red), C3S-E G (blue), and C3S-E N (purple)}
	\label{fig:bar_number_events_verification_pv}
\end{figure}

\newpage
\subsection{Analysis}
\label{sec:analysis}

In this section, we evaluate RES drought events under two different scenarios with fixed installed capacities: the current scenario, with 89\% of wind capacity(6.7 GW) and 11\% of PV capacity (0.6 GW), and a projected scenario for 2030, where wind capacity comprises 53\% (9 GW) and PV capacity increases to 47\% (8 GW). Both scenarios are driven by 45 years of ERA5 data. Using the RES drought identification process described earlier (Section~\ref{sec:res_drought}), we first analyse wind and PV droughts separately before presenting the results for combined (wind + PV) RES droughts.

@@

The first part of our analysis examines the annual number of RES drought events for each of the three datasets. For wind results (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting longer than seven days. For PV results (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. Comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal than those for wind, regardless of duration range or dataset.

The bottom panels show the combination of wind and PV using two different capacity scenarios. With the current RES installation (Fig.~\ref{fig:boxplot_number_events}c), the identified RES droughts are similar to those for wind alone, as expected, due to the dominance of installed wind generation. However, in the projected scenario (Fig.~\ref{fig:boxplot_number_events}d), there is a distinct reduction in the number of events across all datasets and durations (26\% for Atlite, 20\% for C3S-E G, 18\% for C3S-E N).

The median, first, and third quartiles for the Atlite dataset are consistently greater than or equal to those of the other two datasets, regardless of the duration range or type of renewable energy considered.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{number_events/boxplot_number_events}
	\caption{Annual number of RES droughts for a)~Wind, b)~PV, and the combination for the c)~current and d)~projected installed capacity for Atlite (red), C3S-E G (blue), and C3S-E N (purple). The x-axis represents duration ranges in days (lower bound included), while the y-axis indicates the annual number of events. The boxes display the first and third quartiles and the median is marked by a black line. The whiskers indicate the 5th and 95th percentiles}
	\label{fig:boxplot_number_events}	
\end{figure}

@@

The RES drought events identified over the 45 year period were used to calculate the return periods for different RES drought durations.

For wind (Fig.~\ref{fig:return_periods}a), the duration of RES droughts increases in a log-linear fashion across the three datasets.

For PV (Fig.~\ref{fig:return_periods}b), the duration of RES droughts follows a log-linear increase for events lasting less than 16 days. Beyond this threshold, there is a sharp rise in RES drought duration for events up to a one-year return period. This sudden jump reflects the effect of winter on PV generation in Ireland, as PV generation does not exceed the CF threshold for a large part of winter. The jump for the Atlite dataset is not as large as that for the two C3S-E datasets, due to differences between the datasets in PV generation around this critical 0.1 CF threshold during winter. After this jump, the increase in drought duration continues linearly relative to the log of time.

Similar to Fig.~\ref{fig:boxplot_number_events}c, the current installed capacity combination (Fig.~\ref{fig:return_periods}c), shows comparable results as the wind panel (Fig.~\ref{fig:return_periods}a). In the projected scenario (Fig.~\ref{fig:return_periods}d) the return periods for RES droughts increase across all durations. For example, the return period for a five-day event increases from around half a year to 21 months in the Atlite dataset, and from around a year and a half to almost three years for the two C3S-E datasets.

>>>>

\section{Introduction}
\label{sec:Intro}

@@

This study aims to quantify the occurrence of RES drought events in Ireland for two energy scenarios. The first represents the current energy mix, predominantly driven by wind, while the second is based on projected energy scenarios, featuring a more balanced share between wind and PV. A uniform methodology is used to identify RES droughts for each of the three datasets, and differences in the frequency, the duration and the seasonality are presented.

The datasets used in this study are detailed in section~\ref{sec:Data}, which describes their sources, characteristics, and relevance for evaluating RES droughts. Section~\ref{sec:Methods} outlines the energy models used to simulate wind and PV generation and provides the methodology for defining and identifying RES drought events, including the thresholds and metrics applied. In section~\ref{sec:Results}, the models are first verified against actual energy data to assess their accuracy, followed by an analysis of RES drought occurrences across the two energy scenarios, highlighting key differences across the datasets and the effects of varying energy capacities. Finally, section~\ref{sec:Conclusion} offers a discussion of the results in the context of energy reliability and future planning, followed by the main conclusions and recommendations for further research.

\section{Data}
\label{sec:Data}

This study uses publicly available datasets to construct and validate the models for estimating the CF of wind and PV energy. The primary data sources include EirGrid, the transmission system operator for the Republic of Ireland, and the ERA5 reanalysis dataset.

\subsection{Wind and PV Generation \& Capacity}
\label{sec:eirgrid}

EirGrid, which operates in the Republic of Ireland, and SONI, the system operator for Northern Ireland, provide detailed datasets on all wind and PV farms across the island of Ireland (Republic of Ireland and Northern Ireland) from 1990 to the present~\cite{eirgrid}. These datasets include critical information such as each farm’s installed capacity, name, and connection date. To enhance the accuracy of this data, the longitude and latitude for each farm were manually determined through online searches.

The spreadsheet available from the EirGrid website contains two key variables: generation and availability values for wind (from 2014 onward) and PV (from 2018 onward). For this study, availability data, which indicates the amount of energy being delivered to the grid, is used. This value is combined with installed capacity to compute the CF, the main metric for assessing renewable energy output.

Detailed data analysis, supported by access to additional data from EirGrid, revealed that the installed capacity alone does not accurately reflect the true capacity used by the grid. Instead, the Maximum Export Capacity (MEC) values from the spreadsheet are used as a basis. To adjust for actual grid conditions, these MEC values are further multiplied by a factor of 1.4, which provides a more realistic representation of the PV generation potential.

@@

C3S has developed a renewable energy dataset for Europe~\cite{dubus2023energy}, using ERA5 atmospheric variables and weather-to-energy models. This dataset provides hourly CF for wind and PV energy from 1979 to the present. The data are available on the same grid as the ERA5 data, which has a horizontal resolution of 0.25°. The time series are also available for download at two aggregated scales: regional (NUTS 2) and national.

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and actual trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: global horizontal irradiance (referred to here as $ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles found in Germany and replicated in European countries~\cite{saintdrenan2018solar}.

\section{Methods}
\label{sec:Methods}

@@

\subsection{C3S-Energy National}
\label{sec:c3se_n}

@@

\subsection{C3S-E Gridded}
\label{sec:c3se_g}

For the second dataset, we use the gridded dataset from C3S-E to create a CF time series which is representative of the actual RES farm distribution. Using the farms' coordinates, we identify the nearest grid point on the C3S-E grid for each farm. We retrieve the CF values from the C3S-E dataset for its corresponding grid point. We then average the CF for each farm using the installed capacity as weights. This process results in a CF time series for each renewable energy source (wind energy and PV energy), which is representative of the location of the RES farms in Ireland.

@@

For wind power estimation, Atlite uses the wind speed at 100 metres ($u_{100}$, $v_{100}$). We provide Atlite with a wind turbine power curve generated by Renewables.ninja: Enercon E112.4500 with 0.30w smoothing~\cite{staffell2016wake} at a hub height of 100 metres, which it uses when converting wind speed to CF. For PV power estimation, Atlite uses four irradiance variables ($ssr$, $ssrd$, $tisr$, and $fdir$) as well as the air temperature ($t2m$). We use a single PV model~\cite{beyer2004pv}, with the Kaneka Hybrid panel option. The azimuth angle is fixed at 180\textdegree (due south) and we select the optimal tilt angle option.

The selection of a specific wind turbine and PV panel characteristics, are further discussed and explained in section \ref{sec:verification}.

@@

This study investigates wind and PV capacity factors (CF) using three different models, resulting in six individual CF time series based on the installed capacities of 2023. In addition to analysing wind and PV generation separately, a combined CF is computed for each model by averaging wind and PV generation, weighted by their installed capacities at the end of 2023 (6.7 GW for wind and 0.6 GW for PV). This results in a total of nine CF time series for the existing scenario, which represents the current distribution of RES installations in Ireland.

Given that PV capacity in Ireland remains relatively low in 2023, and in order to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considers a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These figures are based on the target energy capacity outlined in the roadmap published by EirGrid~\cite{eirgrid2023future}. In this scenario, new PV time series are generated for both the Atlite and C3S-E G models, while the wind CF time series remain the same, as significant changes in wind farm distribution are not expected. From there, the results of these new time series are combined using updated weights which reflect a more balanced scenario (9 GW for wind and 8 GW for PV).

In total, twelve CF time series are analysed in this study, six for individual wind and PV CF (three models for each source) in the existing scenario, and an additional six time series that include the combined CF for both existing and balanced scenarios across the different models.

This approach allows for a comparative analysis between a wind-dominated system (the existing scenario) and a more evenly distributed system (the balanced scenario), assessing how the balance of RES capacity affects the occurrence of RES droughts. The results for these two scenarios are analysed and discussed in the following sections.

\subsection{RES Drought Definition}
\label{sec:res_drought}

In this study, a RES drought event is defined as occurring when the 24-hour moving average of CF remains below a fixed threshold of 0.1 for a period longer than 24 hours. The choice of this threshold is somewhat arbitrary but aligns with similar studies on low renewable energy production \cite{kaspar2019drought, ohba2022drought, mayer2023drought} . By using a 24-hour moving average, fewer but longer-lasting events are captured compared to using the raw CF time series, which can be more sensitive to short-term fluctuations. Although other methods exist for identifying RES droughts, a fixed threshold has been chosen in this study to enable consistent inter-comparison across the three datasets.

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{droughts_methodology}
	\caption{Wind time series of CF (red) and its 24-hour moving average (green) from the 7th to the 15th of July 2021. CF threshold (dashed black). The grey bar shows the period identified as a wind drought under our definition}
	\label{fig:find_res_droughts}
\end{figure}

Fig.~\ref{fig:find_res_droughts} illustrates the moving average identification method. When applying the moving average, this CF time series identifies a single event lasting 161 hours. In contrast, using the raw CF series would have resulted in three separate events, with durations of 36, 35, and 40 hours.

The moving average approach smooths out short-term fluctuations, so that brief periods above the threshold do not interrupt an otherwise continuous low-CF period. This means that a single hour above the threshold does not "break" a drought event if it is surrounded by prolonged low-generation hours. As a result, fewer but longer-lasting drought events are identified, which may better reflect real-world conditions where energy supply constraints persist over extended periods.

\section{Results}
\label{sec:Results}

@@

\subsubsection{Wind Energy}
\label{sec:wind_verification}

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve} a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provide four associated smoothed power curves. The smoothing is done by using a Gaussian filter with different standard deviations that depend on the wind speed. A wind CF time series for Ireland for all wind turbine power curve options and all smoothings is generated. The performance of each wind turbine power curve is then, calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{power_curve}
	\caption{a/ Power curves of the Enercon E112.4500 with a 0.3 smoothing filter used by Atlite (blue) and the Vestas V136/3450 used by C3S-E (orange) b/ Histograms of wind CF for Ireland from Atlite (blue), C3S-E (orange) and EirGrid (shaded)}
	\label{fig:power_curve}
\end{figure}

The smoothing of the wind turbine power curve represents losses due to the wake effect, which is important when modelling wind energy on larger scales. The histogram in Fig.~\ref{fig:power_curve}b shows that the C3S-E power curve tends to underestimate low CF values and overestimate higher ones, whereas the smoothed Atlite power curve more closely follows the recorded wind data from EirGrid.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_wind_contour}
	\caption{Wind CF density plot of the observed EirGrid CF (vertical axes) and modelled (horizontal axes) CF data for the Atlite  (left panel), C3S-E G (central panel) and C3S-E N (right panel) models}
	\label{fig:wind_verification_contour}
\end{figure}

@@

Fig.~\ref{fig:bar_number_events_verification_wind} presents results for the average annual number of wind drought events during the 2014 to 2023 validation period. The figure shows that Atlite shows the best agreement overall with the observed frequency and duration of wind drought events. This pattern is particularly evident for shorter duration events, which are the most frequent.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_wind_bar}
	\caption{Average annual number of wind drought events for EirGrid (black), Atlite (red), C3S-E G (blue), and C3S-E N (purple)}
	\label{fig:bar_number_events_verification_wind}
\end{figure}

\newpage
\subsubsection{PV Energy}
\label{sec:pv_verification}

The C3S-E datasets calculate PV generation multiple times using the same model with varying azimuth and tilt angles. The results are then aggregated based on a statistical distribution of module angles of Germany and then adapted to other European countries. Similarly to the wind CF verification, the Atlite model allows the user to select a given PV panel characteristics. In this study we have chosen to look at the three PV panels offered by in the Atlite model. Following the same methodology as in the previous section, we compare the three available models using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, we selected the Breyer PV panel model \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{verification_pv_contour}
	\caption{PV 2D density plot of the observed EirGrid (vertical axes) and modelled (horizontal axes) CF series for the Atlite  (left panel), C3S-E G (central panel) and C3S-E N (right panel) models}	
	\label{fig:solar_verification_contour}
\end{figure}

Figure \ref{fig:solar_verification_contour} shows that the three datasets have a similar tendency of overestimating the CF compared to EirGrid. The skill scores presented in Table~\ref{tab:pv_skill_scores} indicate that C3S-E G performs best overall, with the lowest RMSE and a high correlation coefficient, suggesting a closer match to observed data. Atlite shows a slightly lower correlation and a higher RMSE, indicating larger deviations. All models show a slight positive bias, with Atlite exhibiting the most overestimation. These results reflect the broader challenges in accurately modeling PV, where variability is hard to capture. Nonetheless, C3S-E G’s lower RMSE suggests it may be more reliable for PV CF modeling in this context.

\begin{table}[!ht]
	\centering
	\begin{tabular}{l|lll|}
	\cline{2-4}
	& \textbf{Atlite} & \textbf{C3S-E G} & \textbf{C3S-E N} \\ \hline
	\multicolumn{1}{|l|}{\textbf{CC}}   & 0.921           & 0.931            & 0.931            \\ \hline
	\multicolumn{1}{|l|}{\textbf{RMSE}} & 0.119           & 0.090            & 0.113            \\ \hline
	\multicolumn{1}{|l|}{\textbf{MBE}}   & 0.046           & 0.027           & 0.021           \\ \hline
	\end{tabular}
	\caption{Skill scores for wind CF for the three datasets compared to EirGrid data}
	\label{tab:pv_skill_scores}
\end{table}

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity. Due to uncertainties in PV generation data during the first few months after each farm is connected, it is difficult to accurately determine the effective capacity. With over 60\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events. While the results are reasonably aligned in some ranges, the inherent challenges in PV data highlight the limitations of this verification.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_pv_bar}
	\caption{Annual number of PV drought events for EirGrid (black), Atlite (red), C3S-E G (blue), and C3S-E N (purple) in 2023}
	\label{fig:bar_number_events_verification_pv}
\end{figure}

\newpage
\subsection{Analysis}
\label{sec:Analysis}

In this section, we evaluate RES drought events under two different scenarios with fixed installed capacities: the existing scenario, with 89\% of wind capacity (6.7 GW) and 11\% of PV capacity (0.6 GW), and a balanced scenario, where wind capacity comprises 53\% (9 GW) and PV capacity increases to 47\% (8 GW). Both scenarios are driven by 45 years of ERA5 data. Using the RES drought identification process described earlier (Section~\ref{sec:res_drought}), we first analyse wind and PV droughts separately before presenting the results for combined (wind + PV) RES droughts.

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets.

For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to PV's typically lower CF compared to wind, especially in a region like Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

The bottom panels show the combination of wind and PV under the  two capacity scenarios. In the existing RES installation scenario (Fig.~\ref{fig:boxplot_number_events}c), the identified RES droughts closely match those for wind alone, which is expected due to the dominance of installed wind capacity. In contrast, the balanced scenario (Fig.~\ref{fig:boxplot_number_events}d) shows a clear reduction in the number of drought events across all datasets and durations, with a decrease of the total number of events of 56\% for Atlite, 52\% for C3S-E G, and 50\% for C3S-E N. This reduction is attributed to the anti-correlation between wind and PV production throughout the year, a phenomenon well-documented in the literature \cite{kaspar2019drought} and discussed in further detail in the later analysis.

The median, first, and third quartiles for the Atlite dataset are consistently greater than or equal to those of the other two datasets, regardless of the duration range or type of renewable energy considered. This difference arises from the wind turbine power curve model used in the C3S-E datasets, which tends to overestimate the wind CF (Fig.~\ref{fig:wind_verification_contour}). As a result, the overall number of RES droughts is underestimated in the C3S-E datasets compared to Atlite.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{droughts_number_events}
	\caption{Annual number of RES droughts for a)~Wind, b)~PV, and the combination for the c)~current and d)~projected installed capacity for Atlite (red), C3S-E G (blue), and C3S-E N (purple). The x-axis represents duration ranges in days (lower bound included), while the y-axis indicates the annual number of events. The boxes display the first and third quartiles and the median is marked by a black line. The whiskers indicate the 5th and 95th percentiles}
	\label{fig:boxplot_number_events}	
\end{figure}

@@

The RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period refers to the estimated average interval between occurrences of events with a specified duration or intensity. Fig.~\ref{fig:return_periods} illustrates the return periods for varying RES drought durations, highlighting how often different drought lengths are likely to occur across the datasets. This analysis provides insight into the frequency and likelihood of prolonged low-generation periods, which is crucial for evaluating the potential impact of RES droughts on energy reliability.

For wind (Fig.~\ref{fig:return_periods}a), the duration of RES droughts increases in a log-linear fashion across the three datasets. The log-linear trend indicates a predictable relationship between drought duration and occurrence, with longer RES droughts becoming exponentially less likely as duration increases.

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves significantly different than the two C3S-E datasets. The Atlite results follow the ones for wind with a log-linear increase, but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets likely arises from how each model handles PV output near the threshold. Atlite appears to hover slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.
